---
title: "Wrapper function `parlMICE`"
author: "Gerko Vink & Rianne Schouten"
date: "Thursday, November 17th, 2016"
output: html_document
header-includes:
- \usepackage{caption}
- \usepackage{natbib}
bibliography: References.bib
---

```{r, include = FALSE}
require(mice)
source("parallelMICE.R")
```

For big datasets or high number of imputations, performing multiple imputation with function `mice` from package `mice` [@Stef2011] might take a long time. As a solution, wrapper function `parlMICE` was created to enable the imputation procedure to be run in parallel. This is done by dividing the imputations over multiple cores (or CPUs), thereby speeding up the process. 

### Gaining time with `parlMICE`

In Figure \ref{fig1}, the processing time is compared between `mice` and `parlMICE`. Here, a dataset with 1000 cases and 4 variables is imputed for different `m` (starting from 6). Besides, multiple imputation with `parlMICE` is performed with 1, 2 and 3 cores. 

\begin{figure}[h]
  \includegraphics[width=\linewidth]{parlMICE.png}
  \caption{Processing time when multiple imputations are performed with `mice` (conventional) and wrapper function `parlMICE` (1, 2 and 3 cores respectively). The data have 1000 cases and 4 variables, with a correlation of 0.5. 80 percent of the cases have one missing value based on MCAR missingness.}
  \label{fig1}
\end{figure}

From Figure \ref{fig1} it becomes apparent that for a small number of imputations, the conventional `mice` function works faster than the wrapper function. This is the case for `m` $\leq$ 35. From that point on, wrapper function `parlMICE` returns the imputations faster.

The right figure of Figure \ref{fig1} shows the processing time for very high number of imputations. After the intersection, the lines follow the same trend: 3 cores perform the work faster than 2 cores, which work faster than 1 core. Obviously, using `parlMICE` with 1 core is similar to function `mice`, as the figure shows with the red and blue lines.

Another test was run with a dataset with 10000 cases and 8 variables. For this dataset, `parlMICE` performed faster than `mice` starting from the first imputation. The time gain can easily reach 75 percent for 10 imputations and more.

### Default settings

We will now dig into the arguments of function `parlMICE`. Easy imputation of an incomplete dataset (say, `nhanes`) can be performed with `parlMICE` in the following way. 

```{r}
imp <- parlMICE(nhanes)
class(imp)
```

The function returns a `mids` object as created by `mice`. In fact, `parlMICE` makes use of function `ibind` to combine the `mids` objects returned by the different cores. Therefore, the `call` of the `mids` object has slightly changed. 

```{r}
imp$call
```

All other parts of the `mids` object are standard.

### Using `mice` arguments

Function parlMICE is able to deal with the conventional `mice` arguments. In order to change the imputation method from its default (predictive mean matching) to, for example, Bayesian linear regression, the `method` argument can be adjusted.

```{r}
imp <- parlMICE(nhanes, method = "norm")
imp$method
```

In `mice`, the number of imputations is specified with argument `m`. In `parlMICE`, it is possible to use this argument as well, but we advise to rather use the arguments `n.core` and `n.imp.core`. The next section discusses these arguments.  

### Arguments `n.core`, `n.imp.core`

With `n.core`, the number of cores (or CPUs) is given. `n.imp.core` defines the number of imputations per core. Subsequently, the total number of imputations equals `n.core` * `n.imp.core`. As a default, `n.core` is specified as the number of available, logical cores minus 1. The default number of imputations per core has been set to 2. 

Subsequently, running the default `parlMICE` function on a computer with, 

```{r}
detectCores()
```

available, logical cores, results to `m = (4 - 1) * 2 = 6` imputations. We can check this by evaluating the `m` shown in the `mids` object. 

```{r}
imp$m
```

### Argument `seed` 

In simulation studies, it is often desired to set a seed to make the results reproducible. In contrast to `mice`, the seed value for `parlMICE` cannot be defined outside the function. This has to do with the different cores running at the same time (for more information we gladly refer to the `parallel` [manual](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf)). 

A seed value can be used in `parlMICE` with argument `seed`: 

```{r}
imp <- parlMICE(nhanes, seed = 123)
```

### Systems other than Windows

Function `parlMICE` calls for function `parLapply` from the `parallel` [package](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf). Although other options are available, we have chosen for `parLapply` because it allows the use of multiple cores on all computers, including a Windows computer. For the cluster, we rely on the default option in the function `makeCluster`, which is `"PSOCK"`. 

On systems other than Windows, cluster type `"FORK"` might be faster (see the [document](http://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/) written by Max Gordon). Therefore, we advise to change the type to `"FORK"` if your computer does not run on Windows. 

```{r, eval = FALSE}
imp <- parlMICE(nhanes, type = "FORK")
```

### References

Gordon, M. (2015). How-to go parallel in R – basics + tips. Available at [http://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/](http://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/)

Manual base-package `Parallel`, available at [https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf)







